---
title: "STA442 Homework3"
author: "SongQi Wang  1003439442"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}

# install.packages('MEMSS')
# install.packages('nlme')
# install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)
# 
# install.packages("Pmisc", repos = "http://r-forge.r-project.org")
# install.packages("Hmisc")
# install.packages("magrittr") 0
# install.packages("dplyr")  

library(mapmisc)
library(magrittr)    
library(dplyr)         
library(rpart)
library(partykit)
library(knitr)
library(tidyverse)
library(faraway)
library(lme4)      
library(SMPracticals)  
library(nlme)
library(INLA)
library(Pmisc)
library(Hmisc)
library(MEMSS)
library(MASS)
library(ggplot2)
library(kableExtra)
library(expm)

```

# CO2

## Introduction

The dataset of Atmoshperic Carbon Dioxide concentrations is from the observatory in Haiwaii, made available by the Scripps CO2 Program at scrippsco2.ucsd.edu. We would like to know if the following events effect the concentration of CO2.

- the OPEC oil embargo which began in October 1973;
- the global economic recessions around 1980-1982;
- the fall of the Berlin wall almost exactly 30 years ago, preceding a dramatic fall in industrial production in the Soviet Union and Eastern Europe
- China joining the WTO on 11 December 2001, which was followed by rapid growth in industrial production;
- the bankruptcy of Lehman Brothers on 15 September 2008, regarded as the symbolic start of the most recent global financial crisis; and
- the signing of the Paris Agreement on 12 December 2015, intended to limit CO2 emissions.

## Method

We used a Generalized Additive Model from the Gamma family with a log link function.
The model is as follows:

$$\begin{split}
Y_{i} &\sim Gamma(\theta)\\
\log (E(Y)) =\beta_0 + cos(2 \pi x_i)\beta_1 + sin(2 \pi x_i)\beta_2 &+cos(4 \pi x_i)\beta_3+sin(4 \pi x_i)\beta_4 + U(t_{i}) + V_i\\
[U_1...U_T]^T &\sim RW2(0,\sigma_U^2)\\
V_i &\sim N(0,\sigma_V^2)\\
\end{split}$$

where:

- $Y_i$ is the concentration of CO2 measured on the specific day which occured $x_i$ years since Jan 1st 1980.
- $cos(2 \pi x_i)$ and $sin(2 \pi x_i)$ represent yearly fluctuations, $cos(4 \pi x_i)$ and $sin(4 \pi x_i)$ represent biyearly fluctuations.
- $U(t)$ is a second-order random walk
- $V_i$ covers independent variation or over-dispersion

We setthe following as penalized complexity prior:

$$\begin{split}
P(\sigma_U > \frac{log(1.01)}{26}) &= 50\%\ \\
P(\sigma_V > 2) &= 50\%\ \\
\end{split}$$


\newpage

## Result










Additionally, we compare the distribution of data to the model we fit, we can see our model fit the data pretty well.



\newpage

# Heat

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}

# INLA Model

heatUrl = "http://pbrown.ca/teaching/appliedstats/data/sableIsland.rds"
heatFile = tempfile(basename(heatUrl))
download.file(heatUrl, heatFile)
x = readRDS(heatFile)
x$month = as.numeric(format(x$Date, "%m"))
xSub = x[x$month %in% 5:10 & !is.na(x$Max.Temp...C.),]
weekValues = seq(min(xSub$Date), ISOdate(2060, 1, 1,
0, 0, 0, tz = "UTC"), by = "7 days")
xSub$week = cut(xSub$Date, weekValues)
xSub$weekIid = xSub$week
xSub$day = as.numeric(difftime(xSub$Date, min(weekValues),
units = "days"))
xSub$cos12 = cos(xSub$day * 2 * pi/365.25)
xSub$sin12 = sin(xSub$day * 2 * pi/365.25)
xSub$cos6 = cos(xSub$day * 2 * 2 * pi/365.25)
xSub$sin6 = sin(xSub$day * 2 * 2 * pi/365.25)
xSub$yearFac = factor(format(xSub$Date, "%Y"))
lmStart = lm(Max.Temp...C. ~ sin12 + cos12 + sin6 +
cos6, data = xSub)
startingValues = c(lmStart$fitted.values, rep(lmStart$coef[1],
nlevels(xSub$week)), rep(0, nlevels(xSub$weekIid) +
nlevels(xSub$yearFac)), lmStart$coef[-1])

mm = get("inla.models", INLA:::inla.get.inlaEnv())
if(class(mm) == 'function') mm = mm()
mm$latent$rw2$min.diff = NULL
assign("inla.models", mm, INLA:::inla.get.inlaEnv())
sableRes = INLA::inla(
Max.Temp...C. ~ 0 + sin12 + cos12 + sin6 + cos6 +
f(week, model='rw2',
constr=FALSE,
prior='pc.prec',
param = c(0.1/(52*100), 0.05)) +
f(weekIid, model='iid',
prior='pc.prec',
param = c(1, 0.5)) +
f(yearFac, model='iid', prior='pc.prec',
param = c(1, 0.5)),
family='T',
control.family = list(
hyper = list(
prec = list(prior='pc.prec', param=c(1, 0.5)),
dof = list(prior='pc.dof', param=c(10, 0.5)))),
control.mode = list(theta = c(-1,2,20,0,1),
x = startingValues, restart=TRUE),
control.compute=list(config = TRUE),
# control.inla = list(strategy='gaussian', int.strategy='eb'),
data = xSub, verbose=TRUE)

mySample = inla.posterior.sample(n = 24, result = sableRes,
num.threads = 8, selection = list(week = seq(1,
nrow(sableRes$summary.random$week))))
length(mySample)
names(mySample[[1]])
weekSample = do.call(cbind, lapply(mySample, function(xx) xx$latent))

```

## Introduction

The IPCC states that:
*Human activities are estimated to have caused approximately 1.0°C of global warming above preindustrial levels, with a likely range of 0.8°C to 1.2°C. Global warming is likely to reach 1.5°C between 2030 and 2052 if it continues to increase at the current rate.*
Also, we have a dataset of daily maximum temperature data recorded on Sable Island, off the coast of Nova Scotia. We would like to check whether the statement is consistent with the data.

## Method

We used Generalized Additive Model from the student - t distribution with a degree of freedom follows a prior distribution. The model is as follows:

$$\begin{split}
Y_{i} &\sim t(n)\\
Y_i = \beta_0 + cos(2 \pi x_i)\beta_1 + sin(2 \pi x_i)\beta_2 &+ cos(4 \pi x_i)\beta_3+sin(4 \pi x_i)\beta_4 + W(t_i) + U_i + V_i + S_i\\
[W_1...W_T]^T &\sim RW2(0,\sigma_W^2)\\
U_i &\sim N(0,\sigma_U^2)\\
V_i &\sim N(0,\sigma_V^2)\\
S_i &\sim N(0,\sigma_S^2)\\
\end{split}$$

where:

- $Y_i$ is the maximum temperature data recorded on the specific day on Sable Island.
- $cos(2 \pi x_i)$ and $sin(2 \pi x_i)$ represent yearly fluctuations, $cos(4 \pi x_i)$ and $sin(4 \pi x_i)$ represent biyearly fluctuations.
- $W(t)$ is a second-order random walk
- $U_i$ is the random effect of weekLid.
- $V_i$ is the random effect of yearFac.
- $S_i$ covers independent variation or over-dispersion

We setthe following as penalized complexity prior:

$$\begin{split}
P(\sigma_W > \frac{0.1}{52*100}) &= 5\%\ \\
P(\sigma_U > 1) &= 50\%\ \\
P(\sigma_V > 1) &= 50\%\ \\
P(\sigma_S > 1) &= 50\%\ \\
P(n < 10) &= 50\%\ \\
\end{split}$$

\newpage

## Result

Once we have the model, we can take random samples from the posterior and plot them.
From the plot, we can see that the temperature nowadays (2019) is approximately 1.0°C higher than preindustrial levels (1900), with a likely range of 0.8°C to 1.2°C. After zeomed in, we can clearly see that the difference will reach 1.5°C between 2030 and 2052 if it continues to increase at the current rate. Therefore, the data from Sable Island is broadly supportive of this statement from the IPCC.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4}

forXaxis2 = ISOdate(seq(1880, 2040, by = 20), 1, 1,
tz = "UTC")

matplot(weekValues[-1], weekSample, type = "l", lty = 1,
col = "#FF000050", xlab = "Time", ylab = "Degrees C°",
# xlim = ISOdate(c(1880,2052), 1, 1, tz = "UTC"),
xaxt = "n", xaxs = "i", main = "Plot of posterior sample")
matlines(weekValues[-1], sableRes$summary.random$week[,
paste0(c(0.5, 0.025, 0.975), "quant")], type = "l", lwd = 1,
lty = c(1, 2, 2), col = "black")
axis(1, forXaxis2, format(forXaxis2, "%Y"))
legend("topleft", bty = "n", lty = c(1,2,1,1,1),
col = c("black", "black", "red", "blue","purple","darkred"), legend = c("mean",
"quantiles", "samples", "2019","2030","2052"))
abline(v = ISOdate(2019, 1, 1, tz = "UTC"), col = "blue")
abline(v = ISOdate(2030, 1, 1, tz = "UTC"), col = "purple")
abline(v = ISOdate(2052, 1, 1, tz = "UTC"), col = "darkred")


matplot(weekValues[-1], weekSample, type = "l", lty = 1,
col = "#FF000050", xlab = "Time", ylab = "Degrees C°",
xlim = ISOdate(c(2000,2060), 1, 1, tz = "UTC"),
xaxt = "n", xaxs = "i", main = "Plot of posterior sample - zoomed")
matlines(weekValues[-1], sableRes$summary.random$week[,
paste0(c(0.5, 0.025, 0.975), "quant")], type = "l", lwd = 1,
lty = c(1, 2, 2), col = "black")
axis(1, forXaxis2, format(forXaxis2, "%Y"))
legend("topleft", bty = "n", lty = c(1,2,1,1,1),
col = c("black", "black", "red", "blue","purple","darkred"), legend = c("mean",
"quantiles", "samples", "2019","2030","2052"))
abline(v = ISOdate(2019, 1, 1, tz = "UTC"), col = "blue")
abline(v = ISOdate(2030, 1, 1, tz = "UTC"), col = "purple")
abline(v = ISOdate(2052, 1, 1, tz = "UTC"), col = "darkred")

```

\newpage

Additionally, we compare the distribution of data to the model we fit, we can see our model fit the data pretty well.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4}
# Plot

temp_not_na = xSub$Max.Temp...C. %>% na.omit()
mean_temp = mean(temp_not_na)
sd_temp = sd(temp_not_na)
xSub$xRescale = (xSub$Max.Temp...C.-mean_temp)/sd_temp

ggplot(xSub, aes(x=xRescale)) +
  geom_bar(binwidth = 0.5, colour="black", fill="white",aes(y=..density..)) +
  stat_function(fun = dt, args = list(df = 10), colour="red",fill="red", geom="ribbon", alpha=0.2,  mapping = aes(ymin=0,ymax=..y..)) + 
  labs(title="T density fit to Heat data") +
                labs(x="Degrees (Normalized)", y="Density")+
  theme(plot.title=element_text(size=14,
                                    hjust=0.5,
                                    lineheight=1.2))

```



