---
title: "STA442 Homework3"
author: "SongQi Wang  1003439442"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}

# install.packages('MEMSS')
# install.packages('nlme')
# install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE)
# 
# install.packages("Pmisc", repos = "http://r-forge.r-project.org")
# install.packages("Hmisc")
# install.packages("magrittr") 0
# install.packages("dplyr")  

library(mapmisc)
library(magrittr)    
library(dplyr)         
library(rpart)
library(partykit)
library(knitr)
library(tidyverse)
library(faraway)
library(lme4)      
library(SMPracticals)  
library(nlme)
library(INLA)
library(Pmisc)
library(Hmisc)
library(MEMSS)
library(MASS)
library(ggplot2)
library(kableExtra)
library(expm)

```

# CO2

## Introduction

The dataset of Atmoshperic Carbon Dioxide concentrations is from the observatory in Haiwaii, made available by the Scripps CO2 Program at scrippsco2.ucsd.edu. We would like to know if the following events effect the concentration of CO2.

- the OPEC oil embargo which began in October 1973;
- the global economic recessions around 1980-1982;
- the fall of the Berlin wall almost exactly 30 years ago, preceding a dramatic fall in industrial production in the Soviet Union and Eastern Europe
- China joining the WTO on 11 December 2001, which was followed by rapid growth in industrial production;
- the bankruptcy of Lehman Brothers on 15 September 2008, regarded as the symbolic start of the most recent global financial crisis; and
- the signing of the Paris Agreement on 12 December 2016, intended to limit CO2 emissions.

## Method

We used a Generalized Additive Model from the Gamma family with a log link function.
The model is as follows:

$$\begin{split}
Y_{i} &\sim Gamma(O_{i}\lambda_{i})\\
\ln (\lambda_{i}) &=X_{i} \beta + U(t_{i}) + V_i\\
[U_1...U_T]^T &\sim RW2(0,\sigma_U^2)\\
V_i &\sim N(0,\sigma_V^2)
\end{split}$$

where:

- 

## Result





```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}

cUrl = paste0("http://scrippsco2.ucsd.edu/assets/data/atmospheric/",
"stations/flask_co2/daily/daily_flask_co2_mlo.csv")
cFile = basename(cUrl)
if (!file.exists(cFile)) download.file(cUrl, cFile)
co2s = read.table(cFile, header = FALSE, sep = ",",
skip = 69, stringsAsFactors = FALSE, col.names = c("day",
"time", "junk1", "junk2", "Nflasks", "quality",
"co2"))
co2s$date = strptime(paste(co2s$day, co2s$time), format = "%Y-%m-%d %H:%M",
tz = "UTC")
# remove low-quality measurements
co2s[co2s$quality >= 1, "co2"] = NA
plot(co2s$date, co2s$co2, log = "y", cex = 0.3, col = "#00000040",
xlab = "time", ylab = "ppm")
plot(co2s[co2s$date > ISOdate(2016, 3, 1, tz = "UTC"),
c("date", "co2")], log = "y", type = "o", xlab = "time",
ylab = "ppm", cex = 0.5)

timeOrigin = ISOdate(1980, 1, 1, 0, 0, 0, tz = "UTC")
co2s$days = as.numeric(difftime(co2s$date, timeOrigin,
units = "days"))
co2s$cos12 = cos(2 * pi * co2s$days/365.25)
co2s$sin12 = sin(2 * pi * co2s$days/365.25)
co2s$cos6 = cos(2 * 2 * pi * co2s$days/365.25)
co2s$sin6 = sin(2 * 2 * pi * co2s$days/365.25)
cLm = lm(co2 ~ days + cos12 + sin12 + cos6 + sin6,
data = co2s)
summary(cLm)$coef[, 1:2]

newX = data.frame(date = seq(ISOdate(1990, 1, 1, 0,
0, 0, tz = "UTC"), by = "1 days", length.out = 365 *
30))
newX$days = as.numeric(difftime(newX$date, timeOrigin,
                                units = "days"))
newX$cos12 = cos(2 * pi * newX$days/365.25)
newX$sin12 = sin(2 * pi * newX$days/365.25)
newX$cos6 = cos(2 * 2 * pi * newX$days/365.25)
newX$sin6 = sin(2 * 2 * pi * newX$days/365.25)

coPred = predict(cLm, newX, se.fit = TRUE)
coPred = data.frame(est = coPred$fit, lower = coPred$fit - 2 * coPred$se.fit, upper = coPred$fit + 2 * coPred$se.fit)

plot(newX$date, coPred$est, type = "l")

matlines(as.numeric(newX$date), coPred[, c("lower",
"upper", "est")], lty = 1, col = c("yellow", "yellow",
"black"))
newX = newX[1:365, ]
newX$days = 0
plot(newX$date, predict(cLm, newX))

# time random effect
timeBreaks = seq(min(co2s$date), ISOdate(2025, 1, 1,
tz = "UTC"), by = "14 days")
timePoints = timeBreaks[-1]
co2s$timeRw2 = as.numeric(cut(co2s$date, timeBreaks))
# derivatives of time random effect
D = Diagonal(length(timePoints)) - bandSparse(length(timePoints),
k = -1)
derivLincomb = inla.make.lincombs(timeRw2 = D[-1, ])
names(derivLincomb) = gsub("^lc", "time", names(derivLincomb))

# seasonal effect
StimeSeason = seq(ISOdate(2009, 9, 1, tz = "UTC"),
ISOdate(2011, 3, 1, tz = "UTC"), len = 1001)
StimeYear = as.numeric(difftime(StimeSeason, timeOrigin,
"days"))/365.35
seasonLincomb = inla.make.lincombs(sin12 = sin(2 *
pi * StimeYear), cos12 = cos(2 * pi * StimeYear),
sin6 = sin(2 * 2 * pi * StimeYear), cos6 = cos(2 *
2 * pi * StimeYear))
names(seasonLincomb) = gsub("^lc", "season", names(seasonLincomb))
# predictions
StimePred = as.numeric(difftime(timePoints, timeOrigin,
units = "days"))/365.35
predLincomb = inla.make.lincombs(timeRw2 = Diagonal(length(timePoints)),
`(Intercept)` = rep(1, length(timePoints)), sin12 = sin(2 *
pi * StimePred), cos12 = cos(2 * pi * StimePred),
sin6 = sin(2 * 2 * pi * StimePred), cos6 = cos(2 *
2 * pi * StimePred))
names(predLincomb) = gsub("^lc", "pred", names(predLincomb))
StimeIndex = seq(1, length(timePoints))
timeOriginIndex = which.min(abs(difftime(timePoints, timeOrigin)))

# INLA MODEL

mm = get("inla.models", INLA:::inla.get.inlaEnv())
if(class(mm) == 'function') mm = mm()
mm$latent$rw2$min.diff = NULL
assign("inla.models", mm, INLA:::inla.get.inlaEnv())

co2res = inla(co2 ~ sin12 + cos12 + sin6 + cos6 +
f(timeRw2, model = 'rw2',
values = StimeIndex,
prior='pc.prec', param = c(log(1.01)/26, 0.5)),
data = co2s, family='gamma', lincomb = c(derivLincomb, seasonLincomb, predLincomb),
control.family = list(hyper=list(prec=list(prior='pc.prec', param=c(2, 0.5)))),
# add this line if your computer has trouble
# control.inla = list(strategy='gaussian', int.strategy='eb'),
verbose=TRUE)

# Pmisc::priorPost(co2res)$summary[,c(4,3,5)]
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Plots

# Random effect of CO2
matplot(timePoints, exp(co2res$summary.random$timeRw2[,
c("0.5quant", "0.025quant", "0.975quant")]), type = "l",
col = "black", lty = c(1, 2, 2), log = "y", xaxt = "n",
xlab = "Time", ylab = "ppm", main = "Random effect of CO2",
xlim = ISOdate(c(1960,2025), 1, 1, tz = "UTC"))
xaxPred = seq(ISOdate(1950, 1, 1, tz = "UTC"), by = "10 years",
len = 20)
axis(1, xaxPred, format(xaxPred, "%Y"))
legend("topleft", bty = "n", lty = c(1,1,1,1,1,1), 
col = c("red", "orange", "yellow", "green","blue","purple"), legend = c("1973.Oct",
"1980-82", "1989.Nov", "2001.Dec","2008.Sep","2015.Dec", "Now"))
abline(v = ISOdate(1973, 10, 1, tz = "UTC"), col = "red")
abline(v = ISOdate(1980, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1982, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1989, 11, 9, tz = "UTC"), col = "yellow")
abline(v = ISOdate(2001, 12, 11, tz = "UTC"), col = "green")
abline(v = ISOdate(2008, 9, 15, tz = "UTC"), col = "blue")
abline(v = ISOdate(2015, 12, 12, tz = "UTC"), col = "purple")
abline(v = ISOdate(2019, 11, 13, tz = "UTC"), col = "black")

# Estimated smoothed trend of CO2
matplot(timePoints, exp(timePred), type = "l", col = "black",
lty = c(1, 2, 2), log = "y", main = "Estimated smoothed trend of CO2",
xaxs = "i", xaxt = "n", xlab = "Time", ylab = "CO2 (ppm)",
xlim = ISOdate(c(1957,2025), 1, 1, tz = "UTC"))
xaxPred = seq(ISOdate(1950, 1, 1, tz = "UTC"), by = "10 years",
len = 20)
axis(1, xaxPred, format(xaxPred, "%Y"))
legend("topleft", bty = "n", lty = c(1,1,1,1,1,1), 
col = c("red", "orange", "yellow", "green","blue","purple"), legend = c("1973.Oct",
"1980-82", "1989.Nov", "2001.Dec","2008.Sep","2015.Dec", "Now"))B
abline(v = ISOdate(1973, 10, 1, tz = "UTC"), col = "red")
abline(v = ISOdate(1980, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1982, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1989, 11, 9, tz = "UTC"), col = "yellow")
abline(v = ISOdate(2001, 12, 11, tz = "UTC"), col = "green")
abline(v = ISOdate(2008, 9, 15, tz = "UTC"), col = "blue")
abline(v = ISOdate(2015, 12, 12, tz = "UTC"), col = "purple")
abline(v = ISOdate(2019, 11, 13, tz = "UTC"), col = "black")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}


# The derivative of the log concentration of CO2
scaleTo10Years = (10 * 365.25/as.numeric(diff(timePoints,
units = "days")))
derivPred = co2res$summary.lincomb.derived[grep("time",
rownames(co2res$summary.lincomb.derived)), c("0.5quant",
"0.025quant", "0.975quant")]
matplot(timePoints[-1], scaleTo10Years * derivPred,
type = "l", col = "black", lty = c(1, 2, 2), ylim = c(0,
0.1), xlim = range(as.numeric(co2s$date)),
xaxs = "i", xaxt = "n", xlab = "Time", ylab = "Log ppm, change per 10 years",
main = "The derivative of the log concentration of CO2")
xaxPred = seq(ISOdate(1950, 1, 1, tz = "UTC"), by = "10 years",
len = 20)
axis(1, xaxPred, format(xaxPred, "%Y"))
legend("topleft", bty = "n", lty = c(1,1,1,1,1,1), 
col = c("red", "orange", "yellow", "green","blue","purple"), legend = c("1973.Oct",
"1980-82", "1989.Nov", "2001.Dec","2008.Sep","2015.Dec"))
abline(v = ISOdate(1973, 10, 1, tz = "UTC"), col = "red")
abline(v = ISOdate(1980, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1982, 1, 1, tz = "UTC"), col = "orange")
abline(v = ISOdate(1989, 11, 9, tz = "UTC"), col = "yellow")
abline(v = ISOdate(2001, 12, 11, tz = "UTC"), col = "green")
abline(v = ISOdate(2008, 9, 15, tz = "UTC"), col = "blue")
abline(v = ISOdate(2015, 12, 12, tz = "UTC"), col = "purple")
```

From the 2 plots of Estimated smoothed trend and random effect of CO2, we can see the trend is slightly shallower at 1973, between 1980-82, 1989,and 2001. It is actually vary hard to tell. However, we can see how the trend changes vary clearly from the derivative. 

- In 1973, right after the OPEC oil embargo, the slope of CO2 increased, since oil is exteremely important to industrial production. The embargo of oil caused more CO2 emissions.
- Between 1980-1982, the global economic recessions slowed down CO2 emissions.
- The fall of the Berlin Wall, on 9 November 1989, was a pivotal event in world history which marked the falling of the Iron Curtain. Therefore, we can see the slope crashed at the beginning of 1990, marking the dramatic fall in industrial production in the Soviet Union and Eastern Europe.
- On 11 December 2001, China joined the WTO. CO2 emissions increased due to the followed rapid growth in industrial production.
- On 15 September 2008, Lehman Brothers bankrupted, regarded as the symbolic start of the most
recent global financial crisis. But this event seems did not affect the CO2 emissions very much. The level of emission remains the same level around the global financial crisis and it increased in 2009, which reperesnts the end of the crisis.
- on 12 December 2015, the signing of the Paris Agreement intended to limit CO2 emissions. It looked effective, since the emissions of CO2 decreased at the beginning of 2016.




```{r, echo=FALSE, message=FALSE, warning=FALSE}

matplot(StimeSeason, exp(co2res$summary.lincomb.derived[grep("season",
rownames(co2res$summary.lincomb.derived)), c("0.5quant",
"0.025quant", "0.975quant")]), type = "l", col = "black",
lty = c(1, 2, 2), log = "y", xaxs = "i", xaxt = "n",
xlab = "time", ylab = "relative ppm")
xaxSeason = seq(ISOdate(2009, 9, 1, tz = "UTC"), by = "2 months",
len = 20)
axis(1, xaxSeason, format(xaxSeason, "%b"))
timePred = co2res$summary.lincomb.derived[grep("pred",
rownames(co2res$summary.lincomb.derived)), c("0.5quant",
"0.025quant", "0.975quant")]


matplot(timePoints, exp(timePred), type = "l", col = "black",
lty = c(1, 2, 2), log = "y", xlim = ISOdate(c(2010, 
2025), 1, 1, tz = "UTC"), ylim = c(390, 435),
xaxs = "i", xaxt = "n", xlab = "time", ylab = "ppm")
xaxPred = seq(ISOdate(2010, 1, 1, tz = "UTC"), by = "5 years",
len = 20)
axis(1, xaxPred, format(xaxPred, "%Y"))
```

\newpage

Additionally, we compare the distribution of data to the model we fit, we can see our model fit the data pretty well.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Plot

co2_not_na = co2s$co2 %>% na.omit()
EY = mean(co2_not_na)
VarY = (sd(co2_not_na))^2

scale = VarY/EY
shape = EY/scale

ggplot(co2s, aes(x=co2)) +
  geom_bar(binwidth = 2.5, colour="black", fill="white",aes(y=..density..)) +
  stat_function(fun=dgamma, args = list(shape = shape, scale = scale), colour="red",fill="red", geom="ribbon", alpha=0.2,  mapping = aes(ymin=0,ymax=..y..)) + 
  labs(title="Gamma density fit to CO2 data") +
                labs(x="CO2", y="Density")+
  theme(plot.title=element_text(size=13,
                                    hjust=0.5,
                                    lineheight=1.2))


co2res$priorPost = Pmisc::priorPost(co2res)

```



